{RandomRationalPoints} package for \emph{Macaulay2}

### 2021-03-17 FastLinAlg/Paper/RandomRationalPoints.tex
PaperModified To Incorporate dimViaBezout and MultiplicationTable

   \vspace{1em}
    \noindent \textbf{Acknowledgements:} The authors would like to thank David Eisenbud and Mike Stillman for useful conversations and comments on the development of this package.  The authors began work on this package at the virtual Cleveland 2020 Macaulay2 workshop.

@@

    \begin{description}%[]
        \setlength{\itemsep}{5pt}
    \item[\tt Strategy => $\bullet$]   Here the $\bullet$ can be {\tt Default, BruteForce, LinearIntersection, \\ GenericProjection} or {\tt HybridProjectionIntersection}.

@@

    \item {\tt HybridProjectionIntersection} does a generic projection, followed by a linear intersection. 		
    \end{itemize}

    Notice that the speed, or success, varies depending on the strategy.

@@

    When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Even for some simple examples, it could not provide any rational points if the number of trials is not enough. Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly.  A quick comment, while in many examples generic projection works well, there are some examples where it performs quite slowly.  Setting {\tt ProjectionAttempts=>0} in the {\tt Default} strategy is a way to avoid this situation.

@@

o11 = {{31, -15, 53, 41, 87, 9, -21, -43, -12, 85}}      
    \end{verbatim}
\end{quote}\vspace{-1em}
        }}
    \end{example}

@@

\section{A helper function: {\tt randomCoordinateChange}}{\label{helper}}

% We start {\tt randomCoordinateChange}, a helper function that also might be useful to others.

% \item {\tt Codimension}: It takes input an integer. Various functions need to know the codimension/height of the scheme/ideal it is working with.  Setting this to be an integer will tell the function not to compute the codimension and to use this value instead.  The default value is {\tt null}, in which case the function will compute the codimension.  
%%%   
%%   Codimension is not a function, it is an option.
%%%   
\subsection*{\tt randomCoordinateChange:}

@@

\end{example}
\section{Other helper functions: {\tt genericProjection}, {\tt projectionToHypersurface}}\label{projectionfunctions}

@@

		o6 = ({-2, 1, 1}, {0, 1}, {0, 1}, {1} | x2+xy+2y2+2xz-2yz-2z2   -2y-z |)
		{1} | -2x2-xy-2y2-2xz+2yz+2z2 -2x   |
\end{verbatim}
\end{quote}\vspace{-0.5em}    
    }}
\end{example}

>>>>

   \vspace{1em}
    \noindent \textbf{Acknowledgements:} The authors would like to thank David Eisenbud and Mike Stillman for useful conversations and comments on the development of this package.  The authors began work on this package at the virtual Cleveland 2020 Macaulay2 workshop. The authors are also grateful to the reviewers for suggesting and providing preliminary codes to speed up computations, thereby improving the efficacy of the package substantially.

@@

    \begin{description}%[]
        \setlength{\itemsep}{5pt}
    \item[\tt Strategy => $\bullet$]   Here the $\bullet$ can be {\tt Default, BruteForce, LinearIntersection, \\ GenericProjection}, {\tt HybridProjectionIntersection} or {\tt MultiplicationTable}.

@@

    \item {\tt HybridProjectionIntersection} does a generic projection, followed by a linear intersection.

    \item {\tt MultiplicationTable} works for homogeneous ideals only. It cuts down with a linear space to a zero-dimensional projective scheme and computes how the last two variables act on the quotient ring truncated  
    at the regularity. In case of an absolutely irreducible $I$ over the field, we will find with  high probability a point, 
    since the determinant will have a linear factor in at least 50\% of the cases.
    \end{itemize}

    Notice that the speed, or success, varies depending on the strategy (see also \Cref{helper}).

@@

    When working over very small fields, frequently {\tt BruteForce} is most efficient.  This is not surprising as there may not be many points to check, and a surprisingly large percentage of points can be rational in hypersurfaces and complete intersections.  However, if the field size is larger, {\tt BruteForce} will perform poorly.  Even for some simple examples, it could not provide any rational points if the number of trials is not enough. Other strategies work differently on different examples, and even the same strategy can sometimes work very quickly even if it typically works very slowly. Providing {\tt codimension} speeds up computations significantly as computing codimension using the default command {\tt codim} creates a bottleneck. If codimension needs to be computed, we  circumvent this issue using a probabilistic code to compute it (see \Cref{helper}). A quick comment, while in many examples generic projection works well, there are some examples where it performs quite slowly.  Setting {\tt ProjectionAttempts=>0} in the {\tt Default} strategy is a way to avoid this situation primarily.

@@

o11 = {{31, -15, 53, 41, 87, 9, -21, -43, -12, 85}}      
    \end{verbatim}
\end{quote}\vspace{-1em}
        }}
    \end{example}
To enable speed ups as indicated in the above discussion, the {\tt ProjectionAttempts} has been set to $0$ in the {\tt Default} strategy so that {\tt GenericProjection} is not opted for unless specified by user.

@@

\section{Two helper functions: {\tt dimViaBezout} and {\tt randomCoordinateChange}}{\label{helper}}

\subsection*{\tt dimViaBezout:} We thank the reviewer for pointing out that in most of the computations, computing the codimension significantly slows down speed. Instead, we compute the dimension probabilistically using {\tt dimViaBezout}. This function takes as input an ideal $I$ in a polynomial ring over a field and intersects $V(I)$ with successively higher dimensional random linear spaces until there is an intersection.  For example, if $V(I)$ intersect a random line has a point, then we expect that $V(I)$ contains a hypersurface.  If there was no intersection, this function tries a $2$-dimensional linear space, and so on.  This speeds up many computations. The function also takes in optional inputs as described below:

\begin{itemize}
	\item {\tt DimensionIntersectionAttempts}: 
	This integer input denotes the number of linear spaces to try before moving to the next dimension.
	\item {\tt MinimumFieldSize}:
	If the ambient field is smaller than this integer value, it will automatically be replaced with an extension field. The user may set the {\tt MinimumFieldSize} to ensure that the field being worked over is big enough.  For instance, there are relatively few linear spaces over a field of characteristic $2$, and this can cause incorrect results to be provided. 
\end{itemize}

\begin{example} We illustrate the speed difference in this example.
	{\color{blue}{
			\begin{quote}
				\begin{verbatim}
				i3 : kk=ZZ/nextPrime 10^2;

				i4 : S=kk[y_0..y_14];

				i5 : I=minors(2,random(S^3,S^{5:-1}));

				o5 : Ideal of S

				i6 : elapsedTime dimViaBezout(I)
				-- 0.303629 seconds elapsed

				o6 = 7

				i7 : elapsedTime dim I
				-- 1.47916 seconds elapsed

				o7 = 7
				\end{verbatim}
			\end{quote}
	}}
\end{example}

As discussed earlier, one can always compute the codimension (say, $c$) of the input ideal using this function, and provide it as an optional input {\tt Codimension => c} whenever calling {\tt randomPoints}. Alternatively, all the strategies for {\tt randomPoints} (see \Cref{strategydetails}) compute the dimension using the above function, which significantly accelerates the computations.

\subsection*{\tt randomCoordinateChange:}

@@

\end{example}

\section{Other helper functions: {\tt genericProjection}, {\tt projectionToHypersurface}}\label{projectionfunctions}

@@

		o6 = ({-2, 1, 1}, {0, 1}, {0, 1}, {1} | x2+xy+2y2+2xz-2yz-2z2   -2y-z |)
		{1} | -2x2-xy-2y2-2xz+2yz+2z2 -2x   |
\end{verbatim}
\end{quote}  
    }}
\end{example}
